Non-word repetition in Yélî Dnye

### 2020-09-11 manuscript.Rmd
add relevant ethnographic info

### Intro to the people (**mc**) - please feel free to throw away anything that is not useful!

Most speakers of the language grow up speaking it monolingually within their swidden horticulturalist communities until they begin attending school around the age of 7; school instruction is in English. While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission (1953--present) where there appears to be a higher incidence of married-in mothers from other islands [@brownIPchapter], the same region in which the current data were collected. Children in these multilingual families grow up speaking Yélî Dnye and English, Tok Pisin, and/or some other language(s) from the region.

Little is known about language development in children growing up in Rossel Island, a community of primarily subsistence farmers who tend to reside in close-knitted villages where child care is distributed across many individuals, and who typically speak Yélî Dnyé, a phonologically and lexically complex language.

- usually monolingual at home
- schooling in English but it starts at age XX, so not relevant here
- however, some use of English due to immigrants & children of immigrants
- children spend a lot of time with other children
- most parents are subsistence farmers
- parental education generally varies between XX and YY

@@

Naturally, designing the task in this way may render the study of individual variaiton within the population  more difficult because different children are exposed to different items. However, a review of previous work on individual variation suggested to us that many individual differences effects are relatively small, and would not be detectable with the sample size that we could collect in a given visit. <!--M2self: make sure it's apparent how/why a visit is limited-->

>>>>

### The Yélî community
<!--M2A: is this okay?-->
Most speakers of the language grow up speaking it monolingually until they begin attending school around the age of 7; school instruction is in English. While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing]. Children in these multilingual families grow up speaking Yélî Dnye and English, Tok Pisin, and/or some other language(s) from the region.

Most Yélî people are swidden horticulturalists, raising a variety of tubers (e.g., yams, manioc, taro), coconut, banana, and more in their gardens. In addition to garden foods, Yélî children grow up eating fish and foraging for shellfish and nuts, which provide a regular source of extra protein. The typical household in our dataset includes seven individuals and is situated among a collection of 4 or more other households, with structures often arranged around an open grassy area. These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline.

Laying 250 nautical miles off the coast of mainland PNG and surrounded by barrier reef, transport to and from the island is both infrequent and irregular. There is no electricity system (we use solar panels) or motor vehicles on the island, save dinghys with outboard motors that are primarily reserved for medical transport, when there is deisel available. There is a medical radio connection at the health clinic (at the Catholic mission) via which news and messages are often passed along. There is also a cellular tower on the island, though it is typically not functioning. As such, access to news and outside resources, particularly international connections and larger data transfers, is typically unavailable. Our data collection is therefore limited to the duration of the researchers' on-island visits.

Despite restricted outside contact, formal education is a priority for many Yélî families. A recent study surveying more than 40 parents of young children in this region of the island found that nearly all parents completed 6 or more years of education, with 65% of mothers and 49% of fathers completing the full range of locally available education (8 years), with around half of those parents having pursued an additional 2+ years of education on other islands in the region or on mainland PNG [@casillas2020early]. While there are a handful of local schools around the island, these may be well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), and so it is very common for households to billet school children during the weekdays for months at a time (or for whole months if the child's home is very far). Combined with practices of collective child guardianship within their close-knit hamlets, the practical outcome for researchers is that adult consent can often come from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's parents---to this child assent is also culturally pertinent, as child independence is encouraged and respected from toddlerhood onward [@brownIPchildrearing].^[Note that in a few cases this does mean that precise details such as child date of birth, parent age in years, and parent education was only able to be approximated as the parents were residing on another part of the island.] Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently.

While work on Yélî language development is growing [e.g., @brown2011cultural; @brown2014interactional; @brownIPchildrearing; @casillas2020early; @liszkowski2012prelinguistic], our knowledge of children's linguistic development is quite limited, and research into their phonological development in particular has only just begun [e.g., Bram's thesis; MC's CFH data]. What we do know about the early language environment is that, while child-directed speech is relatively infrequent throughout the day [@casillas2020early], shared caregiving practices and a near-universal fondness for social interaction with young children ensure that speech directed to children comes from all types of speakers: women, men, and other children [@bunceURcrosscultural; @casillas2020early]. While speech from women predominates in young children's speech environments on Rossel, as it does elsewhere [@bergelsoncasillas2019what; @bunceURcrosscultural], there is a significant and marked increase in child-directed speech from other children as infants get older [@bunceURcrosscultural; @casillas2020early]. This increase in Child CDS is attributed to the fact that, starting around age 2, children often spend large swaths of the day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 people) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].
<!--M2A: I think a nice way to end this bit would be very briefly (~1-sentence) linking this input scenario to the relevance of looking at NWR (à la Cristia 2020 DevRev). What do you think?-->

@@

Naturally, designing the task in this way may render the study of individual variaiton within the population  more difficult because different children are exposed to different items. However, a review of previous work on individual variation suggested to us that many individual differences effects are relatively small, and would not be detectable with the sample size that we could collect in a given visit.

### 2020-09-11 manuscript.tex
add relevant ethnographic info

\subsubsection{\texorpdfstring{Intro to the people (\textbf{mc}) -
please feel free to throw away anything that is not
useful!}{Intro to the people (mc) - please feel free to throw away anything that is not useful!}}\label{intro-to-the-people-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}

Most speakers of the language grow up speaking it monolingually within
their swidden horticulturalist communities until they begin attending
school around the age of 7; school instruction is in English. While
monolingual Yélî Dnye upbringing is common, multilingual families are
not unusual, particularly in the region around the Catholic Mission
(1953--present) where there appears to be a higher incidence of
married-in mothers from other islands (Brown \& Casillas, in press), the
same region in which the current data were collected. Children in these
multilingual families grow up speaking Yélî Dnye and English, Tok Pisin,
and/or some other language(s) from the region.

Little is known about language development in children growing up in
Rossel Island, a community of primarily subsistence farmers who tend to
reside in close-knitted villages where child care is distributed across
many individuals, and who typically speak Yélî Dnyé, a phonologically
and lexically complex language.

\begin{itemize}
\tightlist
\item
  usually monolingual at home
\item
  schooling in English but it starts at age XX, so not relevant here
\item
  however, some use of English due to immigrants \& children of
  immigrants
\item
  children spend a lot of time with other children
\item
  most parents are subsistence farmers
\item
  parental education generally varies between XX and YY
\end{itemize}

@@

The overall NWR score was M = 61\% (SD = 15\%). Scores based on phonemes
are even higher M = 77\% (SD = 9\%). The phoneme-based normalized
Levenshtein distance was M = 23\% (SD = 9\%), meaning that about a fifth
of phonemes were were substituted, inserted, or deleted. Notice that the
normalized Levenshtein distance is the complement of phoneme-based
scores.

@@

\textbf{Option 1: just numbers, first attempts only } Next, we
investigated patterns of deletion and substitution. Deletions were
relatively rare, with only 19 vowels deleted, and 4 consonants.

As for substitutions, it was as common for a nasal vowel to be produced
as an oral vowel as vice versa (0 oral target vowels produced as nasal
vowels, 0 nasal target vowels produced as oral vowels). Substitutions in
which the oral nature was preserved but the quality of the vowel was
changed were a great deal more common than changes in quality among
nasal vowels (42 oral vowels produced with a different quality; 0 nasal
vowels produced with a different quality). As for consonants,
asymmetries were very marked with more complex consonants (specifically
dptpkpkmknmbghlv) mispronounced as simple ones (specifically
mnlwyvdgptkfhch, 0) than vice versa (0). Simple consonants were
mispronounced as other simple consonants quite frequently (62 simple
consonants mispronounced as other simple ones, compared to 0 complex
ones).

\textbf{Option 2: numbers and proportions, first attempts only} Next, we
investigated patterns of deletion and substitution. Deletions were
relatively rare, with only 19 vowels deleted (about 1.17\% of all vowel
targets), and 4 consonants deleted (about 0.24\% of all consonant
targets).

As for substitutions, it was as common for a nasal vowel to be produced
as an oral vowel as vice versa (0 oral target vowels produced as nasal
vowels or about 0\% out of all oral vowel targets, 0 nasal target vowels
produced as oral vowels or about NA\% out of all nasal vowel targets).
Substitutions in which the oral nature was preserved but the quality of
the vowel was changed were a great deal more common than changes in
quality among nasal vowels (42 oral vowels produced with a different
quality or about 0.29\% out of all oral vowel targets; 0 nasal vowels
produced with a different quality or about NA\% out of all nasal vowel
targets). As for consonants, asymmetries were very marked with more
complex consonants (specifically dptpkpkmknmbghlv) mispronounced as
simple ones (specifically mnlwyvdgptkfhch, 0 times or about NA\% out of
all complex consonant targets) than vice versa (0 times or about NA\%
out of all simple consonant targets). Simple consonants were
mispronounced as other simple consonants quite frequently (62 simple
consonants mispronounced as other simple ones or about 0.30\% out of all
simple consonant targets, compared to 0 complex ones or about NA\% out
of all complex consonant targets).

\textbf{Option 3: numbers and proportions, ALL attempts} Next, we
investigated patterns of deletion and substitution. Deletions were
relatively rare, with only 29 vowels deleted (about 0.55\% of all vowel
targets), and 9 consonants deleted (about 0.26\% of all consonant
targets).

As for substitutions, it was as common for a nasal vowel to be produced
as an oral vowel as vice versa (52 oral target vowels produced as nasal
vowels or about 2.01\% out of all oral vowel targets, 58 nasal target
vowels produced as oral vowels or about 6.46\% out of all nasal vowel
targets). Substitutions in which the oral nature was preserved but the
quality of the vowel was changed were a great deal more common than
changes in quality among nasal vowels (197 oral vowels produced with a
different quality or about 3.20\% out of all oral vowel targets; 23
nasal vowels produced with a different quality or about 7.08\% out of
all nasal vowel targets). As for consonants, asymmetries were very
marked with more complex consonants (specifically dptpkpkmknmbghlv)
mispronounced as simple ones (specifically mnlwyvdgptkfhch, 266 times or
about 5.55\% out of all complex consonant targets) than vice versa (2
times about 5.55\% out of all simple consonant targets). Simple
consonants were mispronounced as other simple consonants quite
frequently (128 simple consonants mispronounced as other simple ones or
about 0.48\% out of all simple consonant targets, compared to 62 complex
ones or about 2.79\% out of all complex consonant targets).

@@

@@

@@

@@

>>>>

\subsubsection{The Yélî community}\label{the-yuxe9luxee-community}

Most speakers of the language grow up speaking it monolingually until
they begin attending school around the age of 7; school instruction is
in English. While monolingual Yélî Dnye upbringing is common,
multilingual families are not unusual, particularly in the region around
the Catholic Mission---the same region in which the current data were
collected---where there is a higher incidence of married-in mothers from
other islands (Brown \& Casillas, in press). Children in these
multilingual families grow up speaking Yélî Dnye and English, Tok Pisin,
and/or some other language(s) from the region.

Most Yélî people are swidden horticulturalists, raising a variety of
tubers (e.g., yams, manioc, taro), coconut, banana, and more in their
gardens. In addition to garden foods, Yélî children grow up eating fish
and foraging for shellfish and nuts, which provide a regular source of
extra protein. The typical household in our dataset includes seven
individuals and is situated among a collection of 4 or more other
households, with structures often arranged around an open grassy area.
These household clusters are organized by patrilocal relation, such that
they typically comprise a set of brothers, their wives and children, and
their mother and father, with neighboring hamlets also typically related
through the patriline.

Laying 250 nautical miles off the coast of mainland PNG and surrounded
by barrier reef, transport to and from the island is both infrequent and
irregular. There is no electricity system (we use solar panels) or motor
vehicles on the island, save dinghys with outboard motors that are
primarily reserved for medical transport, when there is deisel
available. There is a medical radio connection at the health clinic (at
the Catholic mission) via which news and messages are often passed
along. There is also a cellular tower on the island, though it is
typically not functioning. As such, access to news and outside
resources, particularly international connections and larger data
transfers, is typically unavailable. Our data collection is therefore
limited to the duration of the researchers' on-island visits.

Despite restricted outside contact, formal education is a priority for
many Yélî families. A recent study surveying more than 40 parents of
young children in this region of the island found that nearly all
parents completed 6 or more years of education, with 65\% of mothers and
49\% of fathers completing the full range of locally available education
(8 years), with around half of those parents having pursued an
additional 2+ years of education on other islands in the region or on
mainland PNG (Casillas, Brown, \& Levinson, 2020). While there are a
handful of local schools around the island, these may be well out of
walking distance for many children (i.e., more than 1 hour on foot or by
canoe each day), and so it is very common for households to billet
school children during the weekdays for months at a time (or for whole
months if the child's home is very far). Combined with practices of
collective child guardianship within their close-knit hamlets, the
practical outcome for researchers is that adult consent can often come
from a combination of aunts, uncles, adult cousins, and grandparents
standing in for the child's parents---to this child assent is also
culturally pertinent, as child independence is encouraged and respected
from toddlerhood onward (Brown \& Casillas, in press).\footnote{Note
  that in a few cases this does mean that precise details such as child
  date of birth, parent age in years, and parent education was only able
  to be approximated as the parents were residing on another part of the
  island.} Finally, as mentioned, the language of schooling in this
region is English, and so the majority of teens and adults on the island
speak some English---many speak it very fluently.

While work on Yélî language development is growing (e.g., Brown, 2011,
2014; Brown \& Casillas, in press; Casillas et al., 2020; Liszkowski,
Brown, Callaghan, Takada, \& de Vos, 2012), our knowledge of children's
linguistic development is quite limited, and research into their
phonological development in particular has only just begun {[}e.g.,
Bram's thesis; MC's CFH data{]}. What we do know about the early
language environment is that, while child-directed speech is relatively
infrequent throughout the day (Casillas et al., 2020), shared caregiving
practices and a near-universal fondness for social interaction with
young children ensure that speech directed to children comes from all
types of speakers: women, men, and other children (Bunce et al., under
review; Casillas et al., 2020). While speech from women predominates in
young children's speech environments on Rossel, as it does elsewhere
(Bergelson et al., 2019; Bunce et al., under review), there is a
significant and marked increase in child-directed speech from other
children as infants get older (Bunce et al., under review; Casillas et
al., 2020). This increase in Child CDS is attributed to the fact that,
starting around age 2, children often spend large swaths of the day
playing, swimming, and foraging for fruit, nuts, and shellfish in large
(\textasciitilde{}10 people) independent and mixed-age child play groups
(Brown \& Casillas, in press; Casillas et al., 2020).

@@

The overall NWR score was M = 61\% (SD = 15\%). Scores based on phonemes
are even higher M = 78\% (SD = 9\%). The phoneme-based normalized
Levenshtein distance was M = 22\% (SD = 9\%), meaning that about a fifth
of phonemes were were substituted, inserted, or deleted. Notice that the
normalized Levenshtein distance is the complement of phoneme-based
scores.

@@

\textbf{Option 1: just numbers, first attempts only } Next, we
investigated patterns of deletion and substitution. Deletions were
relatively rare, with only 7 vowels deleted, and 3 consonants.

As for substitutions, it was as common for a nasal vowel to be produced
as an oral vowel as vice versa (0 oral target vowels produced as nasal
vowels, 0 nasal target vowels produced as oral vowels). Substitutions in
which the oral nature was preserved but the quality of the vowel was
changed were a great deal more common than changes in quality among
nasal vowels (43 oral vowels produced with a different quality; 0 nasal
vowels produced with a different quality). As for consonants,
asymmetries were very marked with more complex consonants (specifically
dptpkpkmknmbghlv) mispronounced as simple ones (specifically
mnlwyvdgptkfhch, 0) than vice versa (0). Simple consonants were
mispronounced as other simple consonants quite frequently (73 simple
consonants mispronounced as other simple ones, compared to 0 complex
ones).

\textbf{Option 2: numbers and proportions, first attempts only} Next, we
investigated patterns of deletion and substitution. Deletions were
relatively rare, with only 7 vowels deleted (about 0.54\% of all vowel
targets), and 3 consonants deleted (about 0.17\% of all consonant
targets).

As for substitutions, it was as common for a nasal vowel to be produced
as an oral vowel as vice versa (0 oral target vowels produced as nasal
vowels or about 0\% out of all oral vowel targets, 0 nasal target vowels
produced as oral vowels or about NA\% out of all nasal vowel targets).
Substitutions in which the oral nature was preserved but the quality of
the vowel was changed were a great deal more common than changes in
quality among nasal vowels (43 oral vowels produced with a different
quality or about 0.29\% out of all oral vowel targets; 0 nasal vowels
produced with a different quality or about NA\% out of all nasal vowel
targets). As for consonants, asymmetries were very marked with more
complex consonants (specifically dptpkpkmknmbghlv) mispronounced as
simple ones (specifically mnlwyvdgptkfhch, 0 times or about NA\% out of
all complex consonant targets) than vice versa (0 times or about NA\%
out of all simple consonant targets). Simple consonants were
mispronounced as other simple consonants quite frequently (73 simple
consonants mispronounced as other simple ones or about 0.35\% out of all
simple consonant targets, compared to 0 complex ones or about NA\% out
of all complex consonant targets).

\textbf{Option 3: numbers and proportions, ALL attempts} Next, we
investigated patterns of deletion and substitution. Deletions were
relatively rare, with only 11 vowels deleted (about 0.25\% of all vowel
targets), and 8 consonants deleted (about 0.21\% of all consonant
targets).

As for substitutions, it was as common for a nasal vowel to be produced
as an oral vowel as vice versa (52 oral target vowels produced as nasal
vowels or about 2.01\% out of all oral vowel targets, 58 nasal target
vowels produced as oral vowels or about 6.46\% out of all nasal vowel
targets). Substitutions in which the oral nature was preserved but the
quality of the vowel was changed were a great deal more common than
changes in quality among nasal vowels (205 oral vowels produced with a
different quality or about 3.22\% out of all oral vowel targets; 23
nasal vowels produced with a different quality or about 7.08\% out of
all nasal vowel targets). As for consonants, asymmetries were very
marked with more complex consonants (specifically dptpkpkmknmbghlv)
mispronounced as simple ones (specifically mnlwyvdgptkfhch, 250 times or
about 5.57\% out of all complex consonant targets) than vice versa (2
times about 5.57\% out of all simple consonant targets). Simple
consonants were mispronounced as other simple consonants quite
frequently (129 simple consonants mispronounced as other simple ones or
about 1.14\% out of all simple consonant targets, compared to 55 complex
ones or about 2.76\% out of all complex consonant targets).

@@

\hypertarget{ref-bergelsoncasillas2019what}{}
Bergelson, E., Casillas, M., Soderstrom, M., Seidl, A., Warlaumont, A.
S., \& Amatuni, A. (2019). What do North American babies hear? A
large-scale cross-corpus analysis. \emph{Developmental Science},
\emph{22}(1), e12724.
doi:\href{https://doi.org/10.1111/desc.12724}{10.1111/desc.12724}

@@

\hypertarget{ref-brown2011cultural}{}
Brown, P. (2011). The cultural organization of attention. In A. Duranti,
E. Ochs, \& and Bambi B Schieffelin (Eds.), \emph{Handbook of Language
Socialization} (pp. 29--55). Malden, MA: Wiley-Blackwell.

\hypertarget{ref-brown2014interactional}{}
Brown, P. (2014). The interactional context of language learning in
Tzeltal. In I. Arnon, M. Casillas, C. Kurumada, \& B. Estigarribia
(Eds.), \emph{Language in interaction: Studies in honor of Eve V. Clark}
(pp. 51--82). Amsterdam, NL: John Benjamins.

@@

\hypertarget{ref-bunceURcrosscultural}{}
Bunce, J., Soderstrom, M., Bergelson, E., Rosemberg, C., Stein, A.,
Alam, F., \ldots{} Casillas, M. (under review). A cross-cultural
examination of young children's everyday language experiences.

\hypertarget{ref-casillas2020early}{}
Casillas, M., Brown, P., \& Levinson, S. C. (2020). Early language
experience in a papuan community. \emph{Journal of Child Language},
\emph{XX}, XX--XX.

@@

\hypertarget{ref-liszkowski2012prelinguistic}{}
Liszkowski, U., Brown, P., Callaghan, T., Takada, A., \& de Vos, C.
(2012). A prelinguistic gestural universal of human communication.
\emph{Cognitive Science}, \emph{36}(4), 698--713.
doi:\href{https://doi.org/10.1111/j.1551-6709.2011.01228.x}{10.1111/j.1551-6709.2011.01228.x}
